Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-03-03 numerics_paper/version2/numerical_chj.tex
Editorial work on the paper.

\begin{abstract}
The numerical artifact known as a carbuncle in the context of the Euler
equations has been shown to appear also in shallow water simulations.
We study how to deal with carbuncles in the shallow water context,
with an emphasis on a new test problem involving a steady circular hydraulic
jump that can be physically unstable.  We show that numerical methods are prone to
either suppress the instability completely or form carbuncles.
We test existing cures for the carbuncle as well as a new method,
and show results that exhibit the shock instability without unphysical carbuncles.
\end{abstract}

@@

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, the HLLE Riemann solver suppresses the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical}.

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle: it appears when the Roe solver is used, but not when
the HLL solver is used, and can be suppressed by similar numerical methods
\cite{kemm2014note,bader2014carbuncle}.

The most common test problems used to investigate carbuncle formation are
that of bow shock formation or a steady, grid-aligned planar shock.
In both of these problems, the correct behavior is the formation of a stable
shock profile, which is achieved easily by simple methods with excessive
diffusion.  This means these problems are not adequate on their own to
evaluate methods for practical calculations.

Elling \cite{elling2009carbuncle} proposed instead a problem specifically
designed to feature a carbuncle as the physical solution.  This has
been used as a test problem to identify methods that impose excessive dissipation.

Herein we introduce a new and more exacting test problem.  Like some
of the problems above, it includes an equilibrium solution consisting of a
steady shock.  Similar to the Elling problem, the equilibrium is unstable.
However, the correct manifestation of the instability is different from
the carbuncle.  This allows us to distinguish schemes that yield correct
behavior from both those that are too dissipative and those that generate
carbuncles.

In the present work, we provide a test problem that possesses a genuine instability
that leads to carbuncles in many numerical approximations.  This is an ideal
test for assessing numerical methods, since neither the presence of carbuncles
nor the complete absence of instability represents the correct behavior.

@@

Perhaps the first reference to the observation of the circular hydraulic jump
comes from Lord Rayleigh \cite{rayleigh1914theory}, who wrote that it
``may usually be seen whenever a stream of water from a tap strikes a horizontal
surface".  This phenomenon that is familiar in the everyday kitchen sink, is in
fact highly nonlinear and complex.  Near the jet, the flow is shallow and
supercritical, while further away it is deeper and subcritical.  The transition from supercritical
to subcritical flow
occurs in a very narrow region and takes the form of a \emph{jump} or \emph{bore}
that is roughly circular if the surface is flat; we refer to it herein as a
circular hydraulic jump (CHJ).

@@

\subsection{Outline}
In this work, we describe and study the circular hydraulic jump as a new
test problem for shallow water discretizations.  We find that although some
existing methods behave acceptably on previous test problems, they
are not capable of providing accurate solutions for the circular hydraulic jump
across the range of flow regimes we study.  We develop a new entropy-stable
Riemann solver for the shallow water equations that generates solutions with
the correct behavior.

\section{Numerical methods for the shallow water equations}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + f(q)_x + g(q)_y & = 0.
\end{align}

@@

\subsection{Wave propagation methods}\label{sec:waveprop}
Let $Q_i$ represent the average value of the set of
conserved quantities over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(Q^n_{i-1},Q^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the 
Lax-Wendroff-LeVeque (LWL) scheme
\begin{align}\label{second-order_via_fluct}
  Q_i^{n+1} = Q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, 
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align}\label{correction-fluxes}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a TVD-limited version of $\W_{i\pm 1/2}^p$.

We note that, for the conservative Riemann solvers we use herein, the LWL scheme
\eqref{second-order_via_fluct} can also be written purely in flux-differencing
form:
\begin{align}\label{flux-differencing-form}
  Q_i^{n+1}=Q_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right]
 -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
with appropriately chosen first order fluxes $\bfF_{i+1/2}$ and correction
fluxes $\tilde{F}_{i-1/2}$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  Q_i^{n+1} = Q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
 = Q_i^n -\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right].
\end{align}

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  We refer to \cite{ketcheson2020riemann} and references therein for
a full and detailed description of these two Riemann solvers in the context of
the shallow water equations.
Neither of these solvers deals with the carbuncle instability in a fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
Both of these solvers satisfy the conservation property
\begin{align} \label{rs_conservation}
    \amdq_\imh + \apdq_\imh = \bff(\bfq_i) - \bff(\bfq_{i-1}).
\end{align}

@@

Since Roe's solver is not entropy stable, the blended solver is also
not guaranteed to be entropy stable.  We therefore include an additional term
that is chosen to give entropy stability, at least if the second-order correction
fluxes are neglected.  The complete proposed scheme can be written in
flux-differencing form 
We propose
a new scheme consisting of the second-order LWL method \eqref{flux-differencing-form} using
the first-order fluxes
\begin{align} \label{blended-flux-1}
  \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max}\mathbb{I} + (1-\theta_{i+1/2})|\hat A_{i+1/2}| +\lambda_{i+1/2}^{\min}\mathbb{I}\right)(\bfq_{i+1}-\bfq_{i})
\end{align}
and correction fluxes \eqref{correction-fluxes} based on the Roe waves and
the modified speeds
\begin{align} \label{s_p}
    s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p
\end{align}
with
\begin{align}\label{lambda_p}
  \lambda_{i+1/2}^p := \theta_{i+1/2}\lambda_{i+1/2}^{\max} + (1-\theta_{i+1/2})|\hat \lambda_{i+1/2}^p| + \lambda_\iph^{\min}.
\end{align}
Here $\mathbb{I}\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$\lambda_{i+1/2}^{\max}$ is the
upper bound on the wave speed used in Rusanov's Riemann solver (see \S \ref{sec:rusanov}),
and $\hat A_{i+1/2}$ is Roe's averaged flux Jacobian (see \S \ref{sec:roe}).

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$.
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $R_i$ that acts as normalization so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

\clearpage
\section{Numerical results}\label{sec:num}
In this section we present one- and two-dimensional numerical experiments to demonstrate the behavior of the
blended Riemann solver from \S\ref{sec:blended_rs} with the extra entropy dissipation from \S\ref{sec:entropy_stable}.
We compare the behavior of the blended solver against the standard Roe's and Rusanov's solvers.
In most of the experiments we use these Riemann solvers with the LWL method reviewed in \S\ref{sec:waveprop}.
When the exact solution is available, we report a convergence test based on the $L^1$-error for the water height
\begin{align*}
  E_1=\sum_i|K_i|~|h_i-h^e(\bfx_i)|,
\end{align*}
where $|K_i|$ is the length or area of cell $i$,
$h_i$ is the cell average of the water height at cell $i$,
and $h^e(\bfx_i)$ is the exact water height evaluated at the center of cell $i$.
Since the methods under consideration are at most second order accurate, comparison
of cell averages with centered point values is a sufficient way to test their accuracy.
In all experiments we use $g=1$.

@@

\section{Numerical study of the circular hydraulic jump}

@@

\section{Conclusions}
In this work we have introduced a new Riemann solver for the shallow water
equations, described in Section \ref{sec:blended_rs}.  Through numerical tests we have
shown that the solver gives accuracy similar to that of Roe's method and
robustness similar to that of Rusanov's method.  Although the full
second-order method we have proposed is not provably entropy stable or
positivity preserving, it gives improved results for test problems
where both of these properties are important.

We have introduced two new test problems for numerical shallow water solvers,
both consisting of flow in an annulus, with inflow from a jet
at the center and outflow at the outer boundary.  The first test
problem, describe in Section \ref{sec:steady_outflow}, has a smooth solution that can
be computed by solving an ODE and thus serves as a useful test of
accuracy.  The second problem, the circular hydraulic jump, involves
a standing shock wave that can be physically unstable but is
also susceptible numerically to the non-physical carbuncle instability.
Additionally, it involves high-velocity low-depth flow regions where
it is challenging to maintain positivity.
This makes it an excellent problem for testing that schemes are both
robust and not overly dissipative.

>>>>

\begin{abstract}
We investigate the numerical artifact known as a carbuncle, in the solution
of the shallow water equations.  We propose a new Riemann solver that is based
on a local measure of the entropy residual and aims to avoid carbuncles while
maintaining high accuracy.
We propose a new challenging test problem for shallow water codes, 
consisting of a steady circular hydraulic
jump that can be physically unstable.  We show that numerical methods are prone to
either suppress the instability completely or form carbuncles.
We test existing cures for the carbuncle and show that only the proposed
method is able to avoid unphysical carbuncles without suppressing the
physical instability.
\end{abstract}

@@

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and LLF suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical}.

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.

The most common test problems used to investigate carbuncle formation are
that of bow shock formation or a steady, grid-aligned planar shock.
In both of these problems, the correct behavior is the formation of a stable
shock profile without carbuncles.  This is achieved by certain methods
designed specifically to avoid carbuncles, but also by by typical first-order
accurate methods.  Thus these problems are not adequate on their own to
evaluate methods for practical calculations.
Elling \cite{elling2009carbuncle} proposed instead a problem specifically
designed to feature a carbuncle as the physical solution.  This has
been used as a test problem to identify methods that impose excessive dissipation.

Herein we introduce a new and more exacting test problem that arises in a
common physical setting.  Like some
of the problems above, it includes an equilibrium solution consisting of a
steady shock.  Similar to the Elling problem, the equilibrium is unstable.
However, the correct manifestation of the instability is different from
the carbuncle.  This allows us to distinguish schemes that yield correct
behavior from both those that are too dissipative and those that generate
carbuncles.

In the present work, we provide a test problem that possesses a genuine instability
that leads to carbuncles in many numerical approximations.  This is an ideal
test for assessing numerical methods, since neither the presence of carbuncles
nor the complete absence of instability represents the correct behavior.
This test problem is the circular hydraulic jump.

@@

Perhaps the first reference to the observation of the circular hydraulic jump
comes from Lord Rayleigh \cite{rayleigh1914theory}, who wrote that it
``may usually be seen whenever a stream of water from a tap strikes a horizontal
surface''.  This phenomenon that is familiar in the everyday kitchen sink, is in
fact highly nonlinear and unintuitive.  Near the jet, the flow is shallow and
supercritical, while further away it is deeper and subcritical.  The transition from supercritical
to subcritical flow
occurs in a very narrow region and takes the form of a \emph{jump} or \emph{bore}
that is roughly circular if the surface is flat; we refer to it herein as a
circular hydraulic jump (CHJ).

@@

In this work, we describe and study the circular hydraulic jump as a new
test problem for shallow water discretizations.

\subsection{Outline}
In Section \ref{sec2}, we review some existing numerical methods for
the shallow water equations, focusing on certain Riemann solvers.
In Section \ref{sec:blended_rs}, we propose a new Riemann solver that blends
those of Roe and Rusanov in order to avoid carbuncles with being
excessively dissipative.  In Section \ref{sec:num}, we use Clawpack to
compare the performance of the new Riemann solver to existing
solvers on several standard shallow water test problems.  In Section \ref{sec:chj}, we
study the circular hydraulic jump using the newly proposed solver.
We find that although some
existing methods behave acceptably on previous test problems, they
are not capable of providing accurate solutions for the circular hydraulic jump
across the range of flow regimes we study.
Some conclusions and future directions are discussed in Section \ref{sec:conclusion}.

\section{Numerical methods for the shallow water equations} \label{sec2}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + \nabla \cdot f(q)_x & = 0,
\end{align}
where $q=[h, hu, hv]^T$ and the flux functions $f$ is defined in accordance
with \eqref{eq:sw}.

@@

\subsection{Wave propagation methods}\label{sec:waveprop}
Let $Q_i$ represent the average value of the set of
conserved quantities over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(Q^n_{i-1},Q^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the 
Lax-Wendroff-LeVeque (LWL) scheme
\begin{align}\label{second-order_via_fluct}
  Q_i^{n+1} = Q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, 
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align}\label{correction-fluxes}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a limited version of $\W_{i\pm 1/2}^p$, usually
based on a total variation diminishing limiter function.

We note that, for the conservative Riemann solvers we use herein, the LWL scheme
\eqref{second-order_via_fluct} can be written equivalently in flux-differencing
form:
\begin{align}\label{flux-differencing-form}
  Q_i^{n+1}=Q_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right]
 -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
with appropriately chosen first order fluxes $\bfF_{i+1/2}$ and correction
fluxes $\tilde{F}_{i-1/2}$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  Q_i^{n+1} = Q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
 = Q_i^n -\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right].
\end{align}

Next we briefly review two commonly-used Riemann solvers: those of Rusanov and
Roe.  We refer to \cite{ketcheson2020riemann} and references therein for
a detailed description of these two Riemann solvers in the context of
the shallow water equations.
Neither of these solvers deals with the carbuncle instability in a fully
satisfactory way.  Rusanov's solver suppresses the carbuncle but is known to
also suppress related real instabilities, while Roe's solver exhibits carbuncles.
Later we will combine these two solvers in order to better deal with shock instability.
Both of these solvers satisfy the conservation property
\begin{align} \label{rs_conservation}
    \amdq_\imh + \apdq_\imh = \bff(\bfq_i) - \bff(\bfq_{i-1}).
\end{align}

@@

Since Roe's solver is not entropy stable, the blended solver is also
not guaranteed to be entropy stable.  We therefore include an additional term
that is chosen to give entropy stability, at least if the second-order correction
fluxes are neglected.  The complete proposed scheme can be written in
flux-differencing form \eqref{flux-differencing-form} using
the first-order fluxes
\begin{align} \label{blended-flux-1}
  \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max}\mathbb{I} + (1-\theta_{i+1/2})|\hat A_{i+1/2}| +\lambda_{i+1/2}^{\min}\mathbb{I}\right)(\bfq_{i+1}-\bfq_{i})
\end{align}
and correction fluxes \eqref{correction-fluxes} based on the Roe waves with
modified speeds
\begin{align} \label{s_p}
    s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p
\end{align}
where
\begin{align}\label{lambda_p}
  \lambda_{i+1/2}^p := \theta_{i+1/2}\lambda_{i+1/2}^{\max} + (1-\theta_{i+1/2})|\hat \lambda_{i+1/2}^p| + \lambda_\iph^{\min}.
\end{align}
Here $\mathbb{I}\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$\lambda_{i+1/2}^{\max}$ is the
upper bound on the wave speed used in Rusanov's Riemann solver (see \S \ref{sec:rusanov}),
and $\hat A_{i+1/2}$ is Roe's averaged flux Jacobian (see \S \ref{sec:roe}).

@@

\subsection{The entropy residual}
Let $\entvar(\bfq)$ and $\entflux(\bfq)$ denote the entropy variable and entropy flux.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(\bfq) \cdot \left[ \frac{\partial \bfq}{\partial t} + \nabla\cdot \bff(\bfq)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfq)}{\partial t} + \entvar(\bfq) \cdot \nabla\cdot \bff(\bfq)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$.
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfq)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(\bfq) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(\bfq) \cdot \nabla\cdot \bff(\bfq) - \nabla\cdot\entflux(\bfq) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfq$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfq)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(\bfq) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfq)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(\bfq)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

\clearpage
\section{Numerical results}\label{sec:num}
In this section we present one- and two-dimensional numerical experiments to demonstrate the behavior of the
blended Riemann solver from \S\ref{sec:blended_rs} with the extra entropy dissipation from \S\ref{sec:entropy_stable}.
We compare the behavior of the blended solver against the standard Roe's and Rusanov's solvers.
In most of the experiments we use these Riemann solvers with the LWL method reviewed in \S\ref{sec:waveprop}.
When the exact solution is available, we report convergence results based on the $L^1$-error for the water height
\begin{align*}
  E_1=\sum_i|K_i|~|h_i-h^e(\bfx_i)|,
\end{align*}
where $|K_i|$ is the length or area of cell $i$,
$h_i$ is the cell average of the water height at cell $i$,
and $h^e(\bfx_i)$ is the exact water height evaluated at the center of cell $i$.
Since the methods under consideration are at most second order accurate, comparison
of cell averages with centered point values is a sufficient way to test their accuracy.
In all experiments we use $g=1$.

@@

\section{Numerical study of the circular hydraulic jump}\label{sec:chj}

@@

\section{Conclusions}\label{sec:conclusion}
In this work we have introduced a new Riemann solver for the shallow water
equations, described in Section \ref{sec:blended_rs}.  Through numerical tests we have
shown that the solver gives accuracy similar to that of Roe's method and
robustness similar to that of Rusanov's method.  Although the full
second-order method we have proposed is not provably entropy stable or
positivity preserving, it gives improved results for test problems
where both of these properties are important.

We have introduced two new test problems for numerical shallow water solvers,
both consisting of flow in an annulus, with inflow from a jet
at the center and outflow at the outer boundary.  The first test
problem, described in Section \ref{sec:steady_outflow}, has a smooth solution that can
be computed by solving an ODE and thus serves as a useful test of
accuracy.  The second problem, the circular hydraulic jump, involves
a standing shock wave that can be physically unstable but is
also susceptible to the numerical carbuncle instability.
Additionally, it involves high-velocity low-depth flow regions where
it is challenging to maintain positivity.
This makes it an excellent problem for testing that schemes are both
robust and not overly dissipative.
